---
title: "BBC Rap Song - Cleveland plot"
author: "Andrea Scalco"
date: "20/04/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

## Packages

Load packages required.

```{r packages, results='hide', message=FALSE}
# Tidy/wrangle/vis data
library(tidyverse)
# For enhanched texts
library(ggtext)
# To scrape data from wikipedia
library(rvest)
# To clean scraped data
library(stringr)
```

## Data retrieve 

Retrieve data from tidytuesday repository.

```{r data retrieve, results='hide', message=FALSE}
# TidyTuesday (tt) main repo
tt_git_url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020"

# 2020.04.14 datasets
rankings <- readr::read_csv(paste(tt_git_url, "2020-04-14/rankings.csv", sep="/"))
```

## Get wiki data

Collect information about artist activity years from wikipedia. First, create the wiki url of each artist.

```{r prepare wiki urls}
# Create a new col with wiki url
rankings_top10 <- rankings %>%
  top_n(10, points) %>%
  # Replace whitespaces with "_"
  mutate(wiki_artist = str_replace_all(artist, " ", "_")) %>%
  # Replace "&" with "and"
  mutate(wiki_artist = str_replace_all(wiki_artist, "&", "and")) %>%
  # Remove everything after "_ft."
  mutate(wiki_artist = str_remove(wiki_artist, "_ft.(.*)")) %>%
  # Create full link
  mutate(wiki_artist = paste("https://en.wikipedia.org/wiki/", wiki_artist, sep=""))

# Inspect new col
rankings_top10$wiki_artist

# One wrong due to disambiguation, manual edit:
rankings_top10[rankings_top10$artist == "Public Enemy",]$wiki_artist <- "https://en.wikipedia.org/wiki/Public_Enemy_(band)"
```

Second, arrange a function to scrape wikipedia information and get the required information.

```{r}
# Function to get activity years from wiki
get_activity_years <- function(wiki_page) {
  wiki_html <- read_html(wiki_page)
  wiki_data <- html_nodes(wiki_html, "td, th")
  dates <- ""
  for (i in 1:length(wiki_data)) {
    if (str_detect(as.character(wiki_data[i]), "Years")) {
      dates <- as.character(wiki_data[i+1])
      break
    }
  }
  dates <- str_replace_all(dates, "<td>|</td>", "")
  return(toString(dates))
}

# Test function on a single case
get_activity_years("https://en.wikipedia.org/wiki/The_Notorious_B.I.G.")
```

Third, run previous function over wikipedia URLs.

```{r}
# Add a column with the wiki data (use map_chr() from purrr package to iterate)
rankings_top10 <- rankings_top10 %>%
  mutate(activity_years = map_chr(wiki_artist, possibly(get_activity_years, NA_character_)))

# Extract start act year from wiki data
rankings_top10 <- rankings_top10 %>%
  # Get the year (ie the first 4 digits that appears in the data)
  mutate(activity_start = str_extract(activity_years, "[0-9][0-9][0-9][0-9]")) %>%
  # Convert from char to integer
  mutate(activity_start = as.integer(activity_start))
```

## Plot

Plot the artist' debut year and the song's year release. Export plot.

```{r}
# Paste title and artits into one column and covert into fact to preserve the order in the plot
rankings_top10 <- rankings_top10 %>%
  mutate(title_artist = paste(title, artist, sep=" - ")) %>%
  mutate(title_artist = factor(title_artist, levels=rev(title_artist)))

# Define colors to use in the plot
yel1 <- "#FFDF01"
yel2 <- "#EAA21E"
pur1 <- "#560D7F"
blu1 <- "#7a78f0"

# Plot 
rap_plot <- rankings_top10 %>%
  ggplot(aes(y=title_artist)) +
  geom_segment(aes(x=activity_start, xend=year, y=title_artist, yend=title_artist), color="white") +
  geom_point(aes(x=activity_start), col=yel2, size=5) +
  geom_point(aes(x=year), col="royalblue", size=2.5) +
  geom_label(aes(label=ID, x=min(rankings_top10$activity_start)-1.25), size=5, fill=blu1, color="white") +
  scale_x_continuous(limits=c(min(rankings_top10$activity_start)-1.25, max(rankings_top10$year))) +
  theme_minimal() +
  theme(plot.title = element_markdown(color="white"),
        plot.subtitle = element_markdown(color="white"),
        panel.background = element_rect(fill=pur1, color=pur1),
        plot.background = element_rect(fill=pur1),
        axis.line.x = element_line(color="white"),
        axis.line.y = element_blank(),
        axis.text = element_text(color=yel1, face="bold"),
        axis.text.y = element_text(size=10),
        axis.title = element_text(color="white", face="bold"),
        panel.grid.major = element_blank(),
        axis.ticks.x = element_line(color="white")
        ) +
  labs(title = "**<span style='color:#FFDF01'>Artist's debut</span>**
       and **<span style='color:#7a78f0' >song's year release</span>**(Top 10 BBC Music poll)",
       x = "", y = "",
       subtitle = "Sources - Poll by: BBC Music. Dataset by: S. Jockers. Artist debut year: Wikipedia.") 

# Export
ggsave(plot=rap_plot,
       filename="rap_plot.png",
       dpi=150,
       width=275, height=175, unit="mm")

```

